Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene

### 2023-03-01 manuscript/manuscript.Rmd
manuscript and ref corrections

---
title: "A Bayesian model of latitudinal temperature gradients applied to the Early Eocene"

@@

<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->

Abstract
===============================================================================
Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time. However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions. Geological data such as coral reefs, mangroves, and evaporites have the potential to supplement geochemical data by improving spatial coverage and constraining temperature estimates. Historically, these data have been used to qualitatively inform upon palaeoclimatic conditions at broad spatial scales, yet no transfer functions exist to convert geological data into temperature estimates. Quantitative temperature reconstructions have therefore not made use of these data. Likewise, the large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples. Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO. We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data.

@@

Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) provide quantitative palaeotemperature estimates using transfer functions based on experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. 2) Palaeobiological, ecological and lithological proxy data  provide climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012]. 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990]. Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_. For our analyses, this dataset was restricted to the EECO and samples from the continental shelf. Recrystallised $\delta^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999]. This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x). After data filtering, x geochemical proxy samples remained. For a detailed description of each proxy see [@Hollis2019].

@@

Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone. Combining climate-sensitive plant communities [@ @ @], reptiles [@ ] and leaf shapes [@] with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modelling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.

@@

This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies (e.g. @Salonen2019; @Wei2020; @Chandra2021) in a novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures. This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record.

@@

[add table on lat gradients]

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC). In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@ @]

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies. Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

Recent GMSsT estimates of the EECO and of the early Eocene range from 23.4 to 37.1 degC, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020]. Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, most estimates fall in the range of 27 - 29.5 degC [@@@@], similar to our GMSsT estimate of `r eotemp_global$"50%"` degC.

@@

- focus on gradient rather than gmst - gmst can easily be calculated from gradient.

In conclusion, the Bayesian hierarchical model presented in this study represents a significant improvement over existing methods for reconstructing past temperatures by incorporating both geochemical and ecological data. The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction. By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates. The hierarchical structure of the model accounts for the variation of temperature estimates at individual localities and uncertainty associated with the ecological temperature proxies. This improved method is expected to significantly alter the proxy-based temperature reconstructions of the past, particularly for intervals with small or biased samples of temperature proxies.
# Acknowledgements

>>>>

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"

@@

<!-- discarded titles:
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->

Abstract
===============================================================================
Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate. Challenges arise particularly from the patchiness and uneven sampling of palaeoclimate data, and from the disparate nature of the proxy data. Geochemical data usually provide temperature estimates via transfer functions informed by experiments, and transfer functions based on the climatic requirements of modern taxa exist for some fossil plant assemblages. In contrast, most ecological and lithological data, e.g. coral reefs and evaporites, only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have been based on either geochemical or ecological data, but few approaches exist for combining multiple proxy types into a single, quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. We show that the inclusion of ecological data and adequate prior information can substantially reduce uncertainty in palaeoclimate reconstructions, and allow for unbiased temperature estimates from sparse data.

<!--Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time. However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions. Geological data such as coral reefs, mangroves, and evaporites have the potential to supplement geochemical data by improving spatial coverage and constraining temperature estimates. Historically, these data have been used to qualitatively inform upon palaeoclimatic conditions at broad spatial scales, yet no transfer functions exist to convert geological data into temperature estimates. Quantitative temperature reconstructions have therefore not made use of these data. Likewise, the large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples. Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO. We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data. -->

@@

Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) provide quantitative palaeotemperature estimates using transfer functions based on experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. 2) Palaeobiological, ecological and lithological proxy data  provide climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g. @Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990]. Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^{18}O$, $\Delta_{47}$, Mg/Ca and $TEX_{86}$. For our analyses, this dataset was restricted to the EECO and samples from the continental shelf. Recrystallised $\delta^18^O$ samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x). After data filtering, x geochemical proxy samples remained. For a detailed description of each proxy see [@Hollis2019].

@@

Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Marwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modelling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.

@@

This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures. This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record.

<!-- This could be a neat future manuscript (although time...) so not sure if we state this idea, or just cut the previous paragraph -->

@@

[Table 1: Early Eocene latitudinal temperature gradients from the literature]

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC). In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021]

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMsST. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies. Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

Recent GMsST estimates of the EECO and of the early Eocene range from 23.4 to 37.1 degC, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020]. Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5 degC [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our GMsST estimate of `r eotemp_global$"50%"` degC.

@@

The Bayesian hierarchical model presented herein is able to reconstruct latitudinal gradients from both geochemical and ecological proxy data, whilst reflecting the uncertainty associated with the ecological temperature proxies, and accounts for the variation of multiple temperature estimates at individual localities. Using informative prior information allows for accurate temperature reconstructions from records with geographically incomplete sampling. By providing temperature estimates across the entire latitudinal range, this method also facilitates the reconstruction of unbiased global average temperatures. Application of our model to the EECO confirms the existence of latitudinal temperature gradients that were flatter than predicted by most Earth system models and by the majority of proxy-based work. High-latitude pollen records support this interpretation. Our GMsST estimate is in good agreement with most existing estimates, indicating that broadly accurate GMsST reconstructions are possible even with substantial deviations in the shape of the latitudinal temperature gradient. Our new method opens the door for improving the accuracy of proxy-based palaeoclimate reoncstructions and Phanerozoic temperature curves, particularly in intervals with a patchy record, and mitigates the biases incurred from uneven sampling.

<!-- for reconstructing past temperatures by incorporating both geochemical and ecological data. The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction. By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates. The hierarchical structure of the model accounts for the variation of temperature estimates at individual localities and uncertainty associated with the ecological temperature proxies. This improved method is expected to significantly alter the proxy-based temperature reconstructions of the past, particularly for intervals with small or biased samples of temperature proxies. -->

# Acknowledgements
